% Inherit standard global formatting
\input{globals/globals_book.tex}




\begin{document}

\title{Wasteland Tactics}

\begin{titlepage}
\maketitle
\end{titlepage}

\begin{multicols}{2}




\backmatter

\chapter{Introduction}
Wasteland tactics is a tabletop wargame simulating the fast-paced randomness and tactical decision-making of a real battlefield--historical, modern, or in the far future. The game is played between 2 or more players with 28-30mm scale miniatures, 10-sided dice, a tape measure, and a table-sized flat surface preferably covered with evocative terrain and structures!

The rules are fundamentally simple to encourage fast play, yet combine together in deep and interesting ways. The game can be played at any scale: a skirmish between ragtag bands of desperadoes will take under an hour to play, while a company-scale battle with hundreds of soldiers and a dozen monsters and war machines is feasible in under four hours.

These game rules are free and open-source, licensed under the GNU GPLv3 license. This means that you are allowed to copy, redistribute, and modify them, as long as you include a copy of the source code used to generate the final product. Ongoing development takes place at \url{https://github.com/Pointedstick/wasteland-tactics} and you are welcome to participate! Feel free to open Issues for problems you encounter or suggestions for improvement, or submit Pull Requests with proposed changes.


\vfill\null
\columnbreak

\section*{Table of contents}

\toccontents




\chapter{Playing the game}

\section*{Step 1: Choose the game size and opposing forces}
It is considered sportsmanlike to build a balanced force rather than optimizing it to fight any particular enemy.

For a \textbf{small patrol-sized game}, each player has 100 points to build a Battle Group consisting of one Command unit with the \textit{Leadership} rule and any number of models from any other unit—except for those with the \textit{Vehicle} or \textit{Monster} rule, which are not used in games of this size. Each model fights individually as a separate unit and all units with no Suppression Markers on them are scoring units.

For a \textbf{medium platoon-sized game}, each player has 1000 points to build a Battle Group consisting of 1-2 Command units, 3-6 Troops units, 0-2 support units, 0-2 Spearhead units, 0-2 Fast units, 0-2 Heavy units, and 0-1 Transport unit for every non-\textit{Vehicle/Monster} unit taken. In games of this size and above, Troops and Spearhead units with no Suppression Markers on them are scoring units.

For a \textbf{large company-sized game}, each player has 2000 points or more to build one or more Battle Groups, each one chosen like in the medium platoon-sized sized game.




\section*{Step 2: Choose the mission}
Choose one of the following missions, or roll a die to randomly select one. In games where one player is designated the attacker or defender, all other players share the same pool of points to build their Battle Groups.

\subsubsection*{1. Ambush:} Designate one player to be the defender, whose deployment zone is a 10'' wide strip extending from the center of the battlefield to one of the corners. The attackers' deployment zone is everywhere 25'' or greater from the defender's zone. The defender attempts to move units off table edges not touched by their deployment zone and wins a victory if more than half of their units are evacuated in this manner. Otherwise, the attackers share victory.

\subsubsection*{2. Blitz:} Designate one player to be the defender. The attacking players share victory if they have three scoring units entirely within the defender's deployment zone at the end of a round. The defender wins once no attacking players have any scoring units remaining.

\subsubsection*{3. Clear the Landing Zone:} Designate one player to be the defender. Place three obstacle-like pieces of terrain each with a footprint of at least 4'' x 4'' along a line drawn down the center of the board in between the two normal deployment zones. Attackers and defenders both attempt to perform Interact Actions with scoring units that have any models in base contact with the obstacles. Each successful Interact Action made by an attacking unit increases that obstacle's score by 1, while each one made by a defending unit reduces the score by 1. An obstacle's score cannot be negative. Each obstacle that reaches a score of 3 is removed, and the attackers win when all three obstacles have been cleared. The defender wins if 8 rounds have elapsed without this happening; reinforcements are presumed to have arrived and the landing has been foiled!

\subsubsection*{4. Critical Objective:} Place a piece of terrain with a footprint of at least 5'' x 5'' in the center of the board. It is a critical objective that must be controlled at all costs. Each player attempts to perform Interact Actions with scoring units that have any models in base contact with the objective. Each successful Interact Action can either increase that player's score by 1, or reduce another player's score by 2. No player's score may become negative. The first player to reach a score of 5 has succeeded in securing the objective and wins.

\subsubsection*{5. Salvage:} Place five objective markers on the battlefield, each not within any player's deployment zone and at least 5'' away from other objective markers. Scoring units with any models in base contact with an objective may attempt to perform Interact Actions to increase its score; when an objective marker reaches a score of 2, it is removed and given to the player whose unit has secured with it. The first player to control three or more objectives markers wins.

\subsubsection*{6. Evacuation:} Designate one player to be the defender and agree upon a piece of terrain touching the outside of the attackers' deployment zone to be the evacuation point. At the end of each round, each of the defender's units that has the majority of its models within 5'' of the evacuation point unit is removed and yields triple its current (not original) cost in point to the defenders. All other players score points by destroying the defender's units; each completely destroyed unit is worth its cost in points. The player with the most points at the end of the game wins.

\subsubsection*{7. Hold at All Costs:} Place a piece of terrain with a footprint of at least 5'' x 5'' in the center of the board. Designate one player to be the defender, whose deployment zone is a 25'' circle at the center of the battlefield; other players are the attackers. Attackers and defenders both attempt to perform Interact Actions with scoring units that have any models in base contact with the obstacles. Each successful Interact Action made by a defending unit reinforces the objective and increases the score by 1, while one made by an attacking unit reduces the score by 1. The attackers share victory if they ever manage to destroy the objective by reducing its score to -1. The defender wins if 8 rounds have elapsed without this happening; reinforcements are presumed to have arrived and the attack has been foiled!

\subsubsection*{8. Rescue/Capture:} After all players have deployed their forces, they alternate placing ten numbered counters face down not within any player's deployment zone, and at least 5'' away from other objective markers. Then roll a die; the counter showing the rolled number is a key individual in need of rescue (or re-capture). A player may gain (or re-gain) control of the individual by succeeding at an Interact Action with a scoring unit that has any models in base contact with it. Once the individual is under a player's control, it is counted as a unit with a Move characteristic of 6 that cannot be targeted, damaged or destroyed, but may only make Move Actions, and may only do so if still in base contact with a scoring unit. The first player to get the individual to touch the table edge in their deployment zone wins.

\subsubsection*{9. Summon Reinforcements:} For each player, Place one piece of terrain representing communications facilities 10'' away from the center of the table, equidistant from one another. Any player's scoring units with any models in base contact with any of these objectives may succeed at an Interact Action to either increase its player's score by 1, or reduce any other player's score by 2. No player's score may become negative. The first player to reach a score of 8 has succeeded in summoning reinforcements and wins.

\subsubsection*{10. Total Annihilation:} After 8 rounds, each enemy unit destroyed is worth its cost in points, while each unit lost subtracts its cost in points. The player with the most points wins.




\section*{Step 3: Set up the battlefield}
The game may be played anywhere, on any surface of at least 4 square feet, but 6x4 is better. Players take turns placing terrain pieces until a satisfactory battlefield has been assembled that supports the mission type, and each one-square-foot area includes at least one large line-of-sight blocking piece of terrain.

Each piece of terrain should ideally be between 4 and 8 square inches in size, though it's good to include one or two even larger areas as well. Players should come to a mutual agreement regarding the classification of the terrain.




\section*{Step 4: Deploy your forces}
Unless the mission specifies otherwise, there are two deployment zones, each a 10''deep strip emanating from one of the longer board edges, across from one another.

When more than 2 players are playing a mission without attackers and a defender, you may either create additional 10'' deep deployment zones on other board edges, or else some players ally with one another and share a deployment zone and a pool of points to build their Battle Forces.

Each player then determines initiative (see the next section), with a +10 bonus for the defending player in missions with attackers and a defender. The player with the lowest number deploys their Battle Groups first. Then the player with the next highest initiative value places deploys their battle groups, and so on.

In all cases, no unit may be deployed within 25'' from the closest enemy unit.





\section*{Step 5: Start the game}

\subsection*{Phase 1: Determine initiative}

Each player adds up the numbers for all their units' \textit{Leadership} rules. The player with the highest number has the initiative for this round. In the event of a tie, the initiative goes to the player whose Battle Groups contain fewer units.

\subsection*{Phase 2: Activate units}

The player with the initiative selects a unit to activate. Fast, Transport, and \textit{Scouts} units must be activated first, and Heavy units must be activated last. Each unit performs up to two Actions and then deactivates. Units may skip Actions. You may choose a unit's second Action after seeing the results of the first.

A unit with any Suppression Markers loses one Action per Suppression Marker, but may optionally make a Will roll with a penalty equal to its number of Suppression Markers; if the roll is passed, the unit is gripped by extraordinary heroism and ignores the effects of its Suppression Markers during this activation! If the roll is failed, it immediately gains an additional suppression Marker and then deactivates.

After the unit deactivates, the player who rolled the next highest initiative value does the same, and so on. Once all players have activated one unit, the player with the initiative then activates a second unit, and so on. When a player runs out of units to activate, other players continue to activate units until they run out as well.


\subsection*{Phase 3: Resolve morale effects}

After all units on the table have been activated, each unit with any Suppression markers makes a Will roll; if the roll passes, the unit loses one Suppression marker.

After this roll, any unit that still has three or more Suppression markers is immediately destroyed! Its warriors panic and flee, are taken prisoner, or become disorganized to the point of combat ineffectiveness.

Then, any unit with models in base contact with enemy models that lost more models during Fight Actions than it killed during its own Fight Actions against that unit (if applicable) must pass a Will roll or else make a Withdraw Reaction away from the enemy models that it is fighting.

Next, any unit that failed any Life rolls during the current round must make a Will roll, with a -1 penalty if the unit has lost more than half its starting number of models. If the Will roll is failed, the unit gains a Suppression Marker.

After morale effects have been resolved, play proceeds to the next round.

\section*{Achieving victory}

The game ends when one player has achieved the mission's victory condition, or when that condition becomes impossible for any player to achieve. In this case, the game is a draw.




\chapter{Game Mechanics, Models, and Units}

\section*{The golden rules}

More specific rules override more general ones.

If a dispute arises during the course of play, go with whichever option would be the coolest. If this cannot be immediately agreed upon, or if both options are lame, then settle the matter quickly with a game of rock-paper-scissors.




\section*{Models}

Your warriors are represented by painted models. A model's abilities are described by the following characteristics:
\begin{itemize}
    \item \textbf{Move:} Speed across the battlefield in inches.
    \item \textbf{Will:} The die roll needed to perform in stressful situations and cast or resist psychic powers.
    \item \textbf{Life:} The die roll needed to survive being hit by enemy weapons.
    \item \textbf{Weapons:} The unit's weapons.
    \item \textbf{Special Rules:} Special rules permanently affecting all models in the unit.
\end{itemize}

% TODO: mention WYSIWYG or not?




\section*{Units}

A unit may consist of a single model, or multiple models acting as a team. Models in multi-model units must remain within 2'' of at least two other models in the unit (or only one, if the unit consists of only two models). If a unit has any models out of this range at the beginning of a round, it makes a Withdraw Reaction to move them back into range before determining initiative.

When a unit makes a Will roll, use the Will characteristic of the model with the highest one.

Each unit belongs to a certain type:

\begin{itemize}
    \item \textbf{Command:} the Battle Group's leadership
    \item \textbf{Support:} specialized support units
    \item \textbf{Spearhead:} the most elite troops
    \item \textbf{Troops:} regular infantry
    \item \textbf{Fast:} mounted flankers
    \item \textbf{Heavy:} tanks, big guns, and heavy hitters
\end{itemize}




\section*{Rolling dice}

During the game, models will be called upon to ``make a Will roll,'' ``make a Life roll,'' and so on. To do this, roll a ten-sided die (also known as a D10) and compare the number to the characteristic in question. For example, a model with a Will characteristic of 6+ will pass Will rolls on a roll of 6 or higher. For Shoot and Fight rolls, the die roll need to succeed is listed on the relevant weapon itself.

If your D10s show the number 0 instead of 10, treat 0 as 10.

Sometimes, modifiers such as ``+2 Life'' or ``-1 Shoot'' may be applied to the number rolled on the die, making the roll easier or more difficult to pass. Modifiers to die rolls always stack.

If any die roll ever requires a roll of 1+ or less, the result is an automatic success; you may skip rolling. Likewise, any die roll requiring a roll of 11+ or higher is an automatic failure.

If asked to roll a ``D5'', then roll a D10 and divide the number by 2, rounding up. Add modifiers after dividing. For example if rolling ``D5+2'' and the die shows a 9, the D5 roll is a 4, and adding 2 produces a final result of 6.

When rolling dice to resolve the Actions or effects on a whole unit, roll them all at once to save time. However if some models in the unit are affected by different conditions—for example using different weapons, attacking different targets, or struck by weapons with different damage values—the dice should be rolled separately.




\section*{Visibility}

For two units to be visible to one another, at least one model in each unit must be visible to at least one model in the other unit.

A model is considered visible to another model if any part of its base, body, hull, head, or turrets can be seen from the other model's base, body, hull, head, or turrets. Ignore outstretched arms, antennae, gun barrels, wings, etc.

Models in the same unit do not block visibility to one another, but models in other units do—friendly or enemy alike. Gaps between other units' models units also block visibility. Thus a unit blocks visibility to units behind it of equal height or are shorter, but not taller.




\section*{Measuring}

Distances are measured in inches. When measuring the distance between two units, measure between the two closest visible models in those units.

If a model includes a decorative base, measure distances to and from the closest point on the model's base. Otherwise, measure distances to and from the closest point on its body, hull, head, or turrets, ignoring protruding elements as before.






\section*{Base contact}

Models are in base contact if their bases are physically touching. If this is impossible because any of the models do have a base or are standing on elevated pieces of terrain, they count as being in base contact as long as they are as close to each other as physically possible and the higher model is not higher up than the other model is tall.




\section*{Special unit rules}

Special rules with a number at the end stack when combined. For example if a unit has the \textit{Bloodthirsty 1} rule and gains \textit{Bloodthirsty 2} from a Support model, it becomes \textit{Bloodthirsty 3}.

\subsubsection*{Acute Senses:} Models in this unit are equipped with night vision, scanning equipment, or keen natural senses. They ignore Dark/Foggy/Snowing penalties, and \textit{Scout} or \textit{Ambush} units may not be set up within 15'' of any of its models. Additionally, units targeted by this unit do not receive Life bonuses from Defend Actions or a Dig In Reactions.

\subsubsection*{Ambush:} This unit specializes in sneak attack and may optionally not be deployed at the start of the game. Thereafter, any time when it is your turn to activate a unit, you may place this unit anywhere farther than D10 inches away from any enemy models. Then the unit activates with one Action.

If the unit also has the \textit{Scouts} rule, or if any of its models arrive within 5'' of a friendly \textit{Scouts} unit, the unit does not need to be placed farther than D10 inches away from any enemy models.

\subsubsection*{Bloodthirsty X:} Models in this unit are battle-hungry savages and gain a +X bonus to their Move characteristics and Fight rolls during a Fight Action when any of them move into base contact with enemy models and none were already in base contact.

\subsubsection*{Bodyguards X:} Models in this unit are assigned to protect high-value troops or are treated as cannon fodder! They no longer block visibility at all. When a friendly non-\textit{Vehicle/Monster} model in another unit that is in base contact with any models from this unit receive any hits from shooting or melee combat, roll a die for each hit before making Life rolls. On each roll of X or greater, one model from this unit is destroyed, and one hit against the other unit is ignored.

\subsubsection*{Bulky X:} Models in this unit are large. They take up X spaces in \textit{Transport} models and may not take Defend Actions or Go to Ground Reactions.

\subsubsection*{Caster:} This model may cast and nullify psychic powers. All \textit{Caster} models know the following powers automatically:

\begin{itemize}
    \item \textbf{Psychic Fury:} The target unit is hit with a weapon with the characteristics [R20, A1, 6+ D10, Deadly 2]
    \item \textbf{War Blessing:} The target unit receives +1 Shoot and +1 Fight bonuses during its next activation.
\end{itemize}

See the Cast Action for more details.

\subsubsection*{Fearless:} Models in this unit are especially crazed or zealous, and have no regard for their own safety! They ignore Haunted penalties and may not take Defend Actions or Reactions. Whenever the unit would receive a Suppression marker or make a Withdraw Reaction, instead one model in the unit of its owner's choice is destroyed. If all models in the unit are \textit{Tough}, instead one model loses a wound.

\subsubsection*{Hardened:} Models in this unit are armored against anti-personnel weapons and suffer no penalties on their Life rolls against weapons with a Damage value of 1 or 2.

\subsubsection*{Jump/Flying:} Models in this unit are capable of airborne movement. They treat all terrain as Clear Terrain for movement purposes, and measure movement distances in a straight line between the starting and ending locations, ignoring all intervening terrain and obstructions.

If the rule is called ``Flying'', then the models are additionally always considered visible regardless of any terrain. They derive no benefits from Concealment or Cover, may not take Defend Actions or Go to Ground Reactions, and do not block visibility to other units. Enemy models may touch or move over their stand or base as if it were not there.

\subsubsection*{Leadership X:} This unit includes a commander or important officer capable of inspiring your forces. It may issue up to X number of the following Orders per round, at any time, to units in the same Battle Group that have at least one model within 5''. The unit may extend this range to 15'' by passing a Will roll, but the order is wasted if the roll fails.

\begin{itemize}
    \item \textbf{Push forward!} The unit may move an additional D5 inches during its next Move or Fight Action.
    \item \textbf{Strike harder!} You may may re-roll up to 3 dice during the unit's next Shoot or Fight Action.
    \item \textbf{Regroup!} The unit immediately loses a Suppression marker.
    \item \textbf{Seize the initiative!} The unit may be activated next even if it otherwise could not be.
    \item \textbf{Call in an airstrike!} Target a visible enemy unit instead of a friendly unit and roll a die, with a -1 penalty for every full 10'' of distance to the target. On a 3+, the target is attacked by a weapon with the following characteristics: R1000 A4 6+ D4 [Area Effect, Deadly 2, Suppressive 1]. On a roll of 1 or less, the nearest friendly unit becomes the target instead!
\end{itemize}

The same Order may not be issued more than once to the same unit during a battle round.

\subsubsection*{Medic/Repair X:} Models in this unit benefit from the presence of a medic or mechanic. At the end of each round, units with this rule may roll a die; on a roll of X or greater, one destroyed model from the model's own unit or a different friendly unit with any models in base contact is returned to the battlefield with one wound. Alternatively, for \textit{Tough} models that have lost at least one wound, one wound is restored). If the rule is called ``Medic,'' it can only be used on non-\textit{Vehicle} models; if it is called ``Repair,'' it can only be used on \textit{Vehicle} models.

\subsubsection*{Open:} Models in this unit are especially vulnerable to \textit{Area Effect} and \textit{Spray weapons}, which gain \textit{Deadly 2} against them and immediately make an extra free attack against all embarked units.

Units embarked on an \textit{Open} model may issue Orders, take Shoot or Cast Actions (measure distances from the closest point on the transport), and take Fight Actions after disembarking as long as the Open model took fewer than two Move Actions during this round.

\subsubsection*{Pathbound:} Models in this unit become bogged down out of open ground. In difficult terrain, you must choose whether to apply a -6 Move penalty, or else count the area as Dangerous Terrain 9.

\subsubsection*{Pathfinder:} Models in this unit are adept at navigation and survival. They treat Difficult Terrain and vertically Impassible Terrain as Clear Terrain.

\subsubsection*{Price of Failure X:} Models in this unit keep order using the harshest possible methods! When a friendly unit with any models within 5'' of this model fails a Will roll for morale, X number of the closest models from that unit are destroyed. The unit is then considered to have passed the Will roll.

\subsubsection*{Reach:} Models in this unit are especially good at bringing their weapons to bear against the enemy in hand-to-hand combat. In a Fight Action, all models in this unit may strike regardless of how far away they are from enemy models.

\subsubsection*{Scouts:} This unit specializes in infiltration and spotting. It may optionally be deployed last anywhere on the battlefield that leaves all of its models out of sight of all enemy models, and when it provides visibility for friendly \textit{Indirect Fire} weapons, they do not suffer a penalty on their Shoot rolls.

\subsubsection*{Stealth X:} While this unit is in an area of Concealment or Cover, models firing at it suffer an additional -X Shoot penalty.

\subsubsection*{Stubborn X:} Models in this unit are iron-willed and can fight through the most stressful situations undaunted. The unit counts its number of Suppression markers as X lower than it really is.

\subsubsection*{Support X:} This model lends support to comrades. Friendly units that have all models within 10'' of this model and at least one in base contact with it gain all effects and special rules listed in X.

\subsubsection*{Tough X:} Models in this unit are especially huge and durable, and must fail X Life rolls before being destroyed. Use a die to keep track of how many ``wounds'' injured models have remaining.

Individual \textit{Tough} models in multi-model units must be destroyed before other \textit{Tough} models in the unit can be allocated Life rolls.

\subsubsection*{Transport X:} This model is a troop transport and may carry up to X other friendly models. It may contain models from multiple units, and units may begin the game embarked on it.

A unit may embark on a \textit{Transport} model by using a Move Action to place all of its models on top of or within 3'' of it. Embarked units are removed from the battlefield, cannot be interacted with, and may take no Action except to disembark.

Embarked units disembark using a Move Action: instead of moving normally, the unit's models are placed on the battlefield within 3'' of this model. Disembarking units may may not take Fight Actions during the current round. If this model is destroyed, embarked units automatically disembark and gain a Suppression marker.

\subsubsection*{Vehicle/Monster X:} This unit is a vehicle or monster. It can never take Morale tests, make Reactions, accumulate Suppression markers, take Defend Actions, embark on transports, or benefit from Concealment.

It may attack with as many of its weapons as it has remaining wounds, perform Advanced Shooting Tasks for free, and fire \textit{Heavy} weapons after moving. However it may only target units with a majority of their models in front of an imaginary line drawn side-to-side through its center.

The unit may turn or rotate only once (at any point) per Move or Fight Action. The unit and enemy units in base contact with it may use Move Actions to move away from one another without requiring Will rolls, and may take Shoot Actions against one another—however they count as moving if they do so, and cannot use weapons with the \textit{Heavy} or \textit{Area Effect} rule. If enemy units fire upon the unit while any of their other models are in base contact with it, only die rolls of 1 will hit their models instead.

When the unit makes a Fight Action that would bring it into base contact with a non-\textit{Vehicle/Monster} unit, target units may not make a Go to Ground Reaction and are completely destroyed if they make a Counter-Attack Reaction that is not sufficient to destroy the \textit{Vehicle/Monster} unit!

Finally, when a \textit{Vehicle/Monster} model is destroyed, do not remove it; it becomes a wreck/corpse that counts as visibility-blocking Difficult Terrain. And if the name of the rule ends with a number, roll a die when when the model is destroyed; on a roll of X or greater, it explodes or lashes out with its death throes and every model even partially within 4'' (including any embarked models) must pass a Life roll with a -1 penalty or else be destroyed. Then remove the model from the table.

\subsubsection*{Vulnerable X:} This model suffers a -X penalty on Life rolls against Fight Actions and against Shoot Actions from Indirect Fire weapons or units with a majority of models behind an imaginary line drawn side-to-side through its center.




\chapter{Actions and Reactions}

\section*{Move}
Each model in the unit may move on the battlefield a number of inches up to its Move characteristic, and rotate any number of times in any direction.

Vertical pieces of terrain such as walls, fences, and ledges can only be ascended and descended if all of the models in the unit are taller than the distance to be traversed (as before, do not count outstretched arms, wings, weapons, antennas, etc), and models must expend movement distance to do so.

A unit may not use a Move Action to pass through Impassible Terrain or other models, or gaps enemy units' models, or come into base contact with enemy models. Models must end their Move Action in a position where they can stand upright.

A unit must pass a Will roll if it wishes to use a Move Action to move any of its models out of base contact with enemy models. If the roll is failed, the Move Action is wasted.




\section*{Shoot}
\subsection*{1. Choose targets}
A unit with none of its models in base contact with enemy models and armed with at least one ranged weapon may fire its ranged weapons at the closest enemy unit.

The unit must pass a Will roll to perform each of the following Advanced Shooting Tasks:

\begin{itemize}
    \item Target a unit that is not the closest that has multiple models or is a \textit{Vehicle} or \textit{Monster}
    \item Have different models fire at different targets
    \item Extend ranges of non-Autohit weapons by 5'' and suffer a -3 Shoot penalty
\end{itemize}

Each failed roll prevents the unit from performing that Advanced Shooting Task, and the resulting confusion imposes an additional -1 Shoot penalty for the remainder of the Action.

\subsection*{2. Targets may react}
Each targeted unit that has no models in base contact with any friendly units may optionally pass a Will roll to make a Reaction; if the Roll fails, the targeted unit instead gains a Suppression Marker. After any Reactions have been resolved, the Shoot Action ends if no enemy models remain visible and within range.

\subsection*{3. Make Shoot rolls}
For each model in the shooting unit that can see at least one model in its target enemy unit, choose one of its ranged weapons that is within range and make a Shoot roll; the weapon tells you its chance of hitting.

\subsection*{4. Opponent makes Life rolls}
If the roll succeeds, the target unit makes a Life roll with a penalty equal to the Damage characteristic of the weapon used. If the unit contains visible models with differing Life characteristics, the owner allocates Life rolls to individual models before rolling.

\subsection*{5. Remove destroyed models}
For each failed Life roll, a visible model in the unit of its owner's choice is destroyed.

Only models in the targeted unit that are visible to one or more firing models may be removed as casualties; if more kills would be scored than there are visible enemy models, the excess damage is lost.

If any models in the targeted unit are in base contact with models from any of your units, each failed Shoot roll is resolved as a hit against one of those units (the other player decides which!).




\section*{Fight}
\subsection*{1. Choose targets}
A unit armed with at least one melee weapon may take a Fight Action to attack enemy units in hand-to-hand combat. Choose one or more enemy units within the unit's movement range as targets.

\subsection*{2. Targets may react}
Each targeted unit that has no models in base contact with any friendly units may optionally pass a Will roll to make a Reaction; if the Roll fails, the targeted unit instead gains a Suppression Marker. After any Reactions have been resolved, the unit makes a free Move Action during which all of its models are required to move in a straight line directly toward the closest enemy model in a targeted unit. If none of the unit's models end up in base contact with any models in any of the target units, the Fight Action ends.

\subsection*{3. Make Fight rolls}
For each model that is in base contact with either an enemy model or a friendly model in base contact with an enemy model, choose one of the model's melee weapons and make a Fight roll (the weapon tells you its chance of hitting), subject to the following modifiers:

\begin{itemize}
    \item +1 bonus for each Suppression Marker on the enemy unit
    \item +1 bonus if the model's unit outnumbers all target units 2:1 or more
    \item +2 bonus if the model's unit outnumbers all target units 3:1 or more
\end{itemize}

When fighting multiple enemy units, each model's Fight rolls must be allocated to its closest enemy unit with models in base contact. For any model, if this is unclear or all valid targets are equidistant, you may choose its target unit.

\subsection*{4. Opponent makes Life rolls}
For each successful Fight roll, the targeted unit makes a Life roll with a penalty equal to the Damage characteristic of the melee weapon used.

\subsection*{5. Remove destroyed models}
For each failed Life roll, a model of its owner's choice is destroyed, starting with models not in base contact with enemy models. If the unit contains visible models with differing Life characteristics, the owner allocates Life rolls to individual models before rolling.




\section*{Defend}
A unit may take a Defend Action as its last Action. For the duration of the current round, the unit gains a +1 Life bonus against shooting attacks and +1 Will bonus for all purposes except Cast Actions or resisting enemy Cast Actions.

A unit with any Suppression Markers always gains the benefits of the Defend Action, no matter what other Action it takes.

To make it easier to see that the unit has taken a Defend Action, you may turn one of its models on its side.




\section*{Cast}
A unit may take a Cast Action to attempt to manifest a psychic power if at least one of its models has the Caster special rule and none of its models are in base contact with enemy models. One Caster model in the unit chooses a psychic power that it knows and a target unit. The power automatically succeeds.

However, any enemy Caster unit may attempt to nullify the power by passing a Will roll. If the roll passes, the power is nullified. If the power is nullified, the casting unit may attempt to override the nullification; it makes a Will roll of its own, and if that roll succeeds, the nullification is overridden and the power is cast! The nullifying model may then attempt to override that with a further Will roll, and so on. All of these Will rolls have the Unstable 2 rule, and benefit from a +1 bonus for every additional Psychic model in the casting or nullifying unit.

This continues until the power is cast, the caster or nullifier opts to stop making further Will rolls, or one of the models is destroyed by the Unstable rule triggering.

If the power is cast successfully, it unleashes the effects detailed in the power's description.




\section*{Interact}
A scoring unit may take an Interact Action to attempt to secure a mission objective. Choose one of the following:
\begin{itemize}
    \item \textbf{Slow and steady:} Do nothing, but then the unit's second Interact Action during this activation automatically succeeds
    \item \textbf{Go for broke:} Pass a will roll to succeed
\end{itemize}

What happens if the Action succeeds depends on the mission being played.




\section*{Reactions}
At various times, a unit may be permitted or required to make a Reaction. Reactions are always made outside of the normal sequence of activation, regardless of whether the reacting unit had already activated, or how many Actions it has available. The reacting unit gains a Suppression Marker after its Reaction takes effect.



\subsection*{Withdraw}
The unit immediately makes a Move Action away from whatever enemy unit provoked the Reaction. All models in the unit must move the full distance in a straight line, with no maneuvering or routing around obstacles. If this cannot be done because any model would come into base contact with impassible terrain or any enemy models, the unit is destroyed; cohesion and morale collapse after finding the retreat corridor blocked or filled with enemy forces!



\subsection*{Go to Ground}
The unit gains a +2 Life bonus against the Action of the enemy unit it is reacting to.



\subsection*{Counter-Attack}
The unit immediately takes a free Shoot Action against the unit that provoked the Reaction, but with a -1 Shoot penalty. If it is reacting to a Fight Action, all of its ranged weapons are automatically considered to be in range during this free Shoot Action.

This Reaction may only be chosen if the distance between the two units is 15'' away or closer.




\chapter{Weapons}

Every model has one or more weapons, each with four characteristics expressed like so:

\textbf{Pistol}: R10 5+ D1 [Assault]

\begin{itemize}
    \item \textbf{R[number]:} The weapon's range. If present, it is a ranged weapon; otherwise it is a melee weapon. If there are two numbers, the first one is a minimum range, measured to the farthest model in the target unit.
    \item \textbf{[number]+:} The weapon's chance of hitting its target. Each Shoot or Fight roll must be equal to or greater than the number to score a hit. ``Autohit'' means it automatically hits!
    \item \textbf{D[number]:} The weapon's damage. If present, subtract this number from Life rolls made by models struck with this weapon.
    \item \textbf{[special rules]:} This weapon's special rules.
\end{itemize}




\section*{Special weapon rules}

\subsubsection*{Assault:} This weapon can be used effectively on the move at close range. It may be used during a Fight Action as though it were a melee weapon, and once per round, it may be fired for free at the end of a Move Action, exactly as with a Shoot Action.

\subsubsection*{Critical X:} This weapon is capable of dealing terrible damage with a lucky hit. If it would automatically hit, roll anyway. For each die that shows an X or higher, the target fails a life roll automatically, with double the value of the weapon's \textit{Deadly} rule, if it has one. If not, it becomes \textit{Deadly 2}.

\subsubsection*{Deadly X:} This weapon is especially damaging, and each Life roll failed to it deals X ``wounds'' to \textit{Tough} models. If it would deal more wounds than the model has left, the excess damage is lost.

\subsubsection*{Indirect Fire:} When firing this weapon, you may opt to suffer a -2 Shoot penalty and be able to target units that are not visible to its firer, but are visible to and have any models within 15'' of a friendly Spearhead or \textit{Scouts} unit.

\subsubsection*{Heavy:} This weapon requires time to set up and cannot be fired if the wielding model (or a \textit{Transport} it is embarked on) moved at all during this round.

\subsubsection*{Rapid/Area Effect/Spray X:} This weapon inflicts damage over a wide area. If the weapon hits, the target unit must take X Life rolls, not just one.

If the rule is called ``Area Effect'' or ``Spray'', it additionally negates Shoot penalties from firing at units in areas of Concealment or Cover, but X is limited to the number of models in the target unit, and it may not be used in melee combat by models that began the round in base contact with enemy models.

If the rule is called ``Spray'', the weapon additionally negates Life bonuses granted by Cover, Defend Actions, and Go to Ground Reactions.

\subsubsection*{Slow:} This weapon is slow to reload and may only be used once per round.

\subsubsection*{Sniper:} This weapon may be used to fire at units that are not the closest visible target—including single-model units—without requiring a Will roll. However the weapon's Damage value is reduced to 0 against Vehicles without the Open rule.

\subsubsection*{Suppressive/Terrifying X:} This weapon's effects are intensely frightening. Once per activation (your choice as to when), units hit with this weapon must immediately pass a Will roll with a -X penalty or else gain a Suppression Marker. If the rule is called ``Terrifying'', a unit failing its Will roll instead makes a Withdraw Reaction.

If the unit is firing multiple \textit{Suppressive} and/or \textit{Terrifying} weapons at the same target, apply only the most severe effects, but increase the penalty on the target unit's Will roll by one for each additional \textit{Suppressive} or \textit{Terrifying} weapon in the firing unit, to a maximum penalty of -5 Will.

\subsubsection*{Unstable X:} This weapon may damage its wielder! If it would automatically hit, roll anyway. For each die that shows an X or lower, the firing model fails a Life roll.




\chapter{Terrain and Cover}

Before a game, the players should classify each area of terrain with the following characteristics. It is permissible for areas of terrain to have multiple classifications; For example a forest would provide both Difficult Terrain and Concealment.




\section*{Clear Terrain}
\textit{For example: open fields, deserts, streets, areas of very sparse vegetation or rubble, hilltops}

No barriers to movement or visibility. Clear Terrain is everything not otherwise classified, as well as horizontal gaps no wider than a model is tall or wide.




\section*{Difficult Terrain}
\textit{For example: infantry going up a hill, cratered ground, ruins, rubble, barricades, sandbag walls, forests, shallow rivers, swamps}

Units receive -2 Move and -2 Shoot penalties during Move or Fight Actions that take any of their models into, out of, or through areas of Difficult Terrain. Units that move through multiple areas of Difficult Terrain use the worst penalties rather than combining them.




\section*{Dangerous Terrain X}
\textit{For example: minefields, quicksand, rivers}

Roll a die for every model in a unit that moves into, out of, through, or comes into base contact with Dangerous Terrain. For each roll of X+ (an appropriate default value is 9), one model is destroyed.




\section*{Impassible Terrain}
\textit{For example: lava floes, deep water, solid windowless walls, cliffs and very steep slopes}

Models may not move into or through Impassible Terrain. If they are forced to do so for any reason, they are immediately destroyed.




\section*{Concealment}
\textit{For example: tall grass, wooden ruins, sparse forests, scrap piles, fences, smoke}

A unit suffers a -2 Shoot penalty when it shoots at a target unit with a majority of its model inside or behind an area of Concealment. This does not apply to targets inside the same area of Concealment as the shooting unit, or to units inside an area of Concealment firing on targets outside of it.

Shots cannot be traced through two areas of Concealment; the second one will block visibility to models behind (but not within) it, up to the height of the tallest element.




\section*{Cover}
\textit{For example: fortifications, metal or masonry ruins, rock piles, trenches, barricades, sandbag walls, being obscured by the top of a slope}

Exactly like Concealment, but a single area blocks visibility to models behind it up to the height of the tallest element and units gain additional protection: the Defend Action's bonuses are doubled, and they gain additional +1 Life bonus against shooting attacks and +1 Will bonus for all purposes except Cast Actions or resisting enemy Cast Actions.


\section*{Buildings and Ruins}
Intact buildings are treated as solid areas of impassible terrain.

Ruins are treated as areas of Concealment or Cover, depending on the method of construction.

Non-\textit{Vehicle/Monster} models may spend movement distance to ascend and descend within ruins that are modeled with multiple levels or floors, even if they are shorter than the distance to be ascended or descended.





\chapter{Advanced Rules}
These optional rules will add additional depth, randomness, and tactical considerations. Feel free to use all, some, or none of them, so long as everyone playing agrees.




\section*{More granular attacks}
For \textit{Rapid} weapons, players may choose to roll X dice instead of rolling one die and multiplying it by X if it hits. This is especially suitable for vehicles and single model units.




\section*{Advantageous firing positions}
When a unit takes a Shoot Action targeting an enemy unit with more than 3 models whose ``footprint'' on the battlefield is at least twice as deep as it is wide from the perspective of the firing unit, it gains a +1 Shoot bonus and the enemy unit receives a -1 Life penalty.

Similarly, when a unit with more than 3 models takes a Shoot Action, if it has a battlefield footprint that is at least twice as wide as it is deep from the perspective of the target unit, it gains a +1 Shoot bonus.

Finally, a unit receives a -1 Life penalty when it is the target of a Shoot Action from a unit in the opposite quadrant from the last unit that fired upon it, and/or from a unit consisting entirely of models standing on terrain that elevates them higher up than the height of the tallest target model in the target unit.

If a unit's footprint or elevation are ambiguous, err on the side of applying no bonuses or penalties.




\section*{Partially concealed units}
When a unit targeted with a Shoot action is benefiting from the effects of Concealment or Cover because only a majority and not all of its models are in the terrain area, the shooting unit may optionally focus fire on only the fully visible models. In this case, those models lose their Concealment or Cover bonuses against this attack, but the unit's other models are no longer considered visible at all.




\section*{Occupying buildings}
Intact buildings can be occupied and are treated as stationary \textit{Transport} models that cannot attack, with the following rules:

\subsubsection*{Light building:} Life 5+, Open, Tough 10, Transport 10, Vehicle, becomes Ruins when destroyed

\subsubsection*{Heavy or fortified building:} Life 3+, Open, Tough 10, Transport 10, Vehicle, becomes Ruins when destroyed

Especially large buildings could instead have Transport 20 or higher.




\section*{Advantages and disadvantages}
Roll a die to determine your Battle Groups' advantages for this battle:
\subsubsection*{1-2 No advantages:} Do not roll for Disadvantages either.

\subsubsection*{3 Heroic figure:} Pick a Command unit; increase the number of its \textit{Leadership} rule by 1, and give it permanent bonuses of +1 Shoot and +1 Fight. If it is a single model, increase the value of its \textit{Tough} rule by 2.

\subsubsection*{4 Good intel:} You always have the initiative and your units suffer no Shoot penalties when firing at enemy units in Concealment or Cover.

\subsubsection*{5-6 Cutting-edge equipment:} The Damage value of all of your units' weapons is increased by 1.

\subsubsection*{7-8 Grizzled veterans:} All units gain the Stubborn 1 rule.

\subsubsection*{9-10 Reinforcements available:} Once per battle when it is your turn to activate a unit, you may bring back a single destroyed Troops unit; place it in its original form anywhere within your deployment zone. It then immediately activates.

\vspace{2em}

...And then roll a die to determine your Battle Groups' disadvantages for this battle:

\subsubsection*{1-2 Exhausted:} All units' Move and Will characteristics are reduced by 1.

\subsubsection*{3-4 Low ammunition:} Anytime a unit makes a shoot Action, if half or more of the dice are failures, the unit's Shoot rolls suffer an additional -1 penalty for the rest of the game. This can happen multiple times!

\subsubsection*{5-6 Terrible commander:} The unit with the highest \textit{Leadership} has the value of its \textit{Leadership} rule reduced to 0.

\subsubsection*{7-8 Desertion:} Roll a die for each of your units one at a time (you may choose the order) until the die shows a 1; that unit has deserted and is not included in this game.

\subsubsection*{9-10 Inexperienced:} Every unit suffers a permanent -1 penalty on its Shoot and Fight rolls.




\section*{Handicap}
To simulate a mismatched or hopeless battle, or to even out differences in player skill, one player may opt to take one or more disadvantages without any advantages,  or allow any opponents to take one or more advantages without any disadvantages.




\section*{Battlefield conditions}
Choose a suitably epic battlefield condition from the following list, or determine randomly by rolling a die:

\subsubsection*{1-4 Normal conditions:} Play the game normally.

\subsubsection*{5 Dark/Foggy/Snowing:} Units that opt to move faster than 5'' during a Move or Fight Action count as being inside Dangerous Terrain 9+, and every unit suffers a -1 Shoot penalty for each full 10'' separating the firing unit and its target.

\subsubsection*{6 Muddy/Swampy/Snowy:} All areas of Open Terrain become Difficult Terrain. Areas that were already Difficult terrain reduce movement by 3.

\subsubsection*{7 Rainy:} Apply the penalties of both the ``Dark/Foggy'' and ``Muddy/Swampy'' conditions.

\subsubsection*{8 Haunted:} All models suffer a permanent -1 Will penalty.

\subsubsection*{9 Corrosive atmosphere:} All models suffer a permanent -1 Life penalty.

\subsubsection*{10 Dangerous ground:} Every time a unit enters a piece of terrain for the first time, roll a die. On a roll of 7+, that area of terrain is Dangerous Terrain 8 for the rest of the game. Mark it accordingly so nobody forgets.




\section*{Random battle events}
At the beginning of each round, the player with the initiative rolls a die. On a roll of 9+, one of the following random events immediately occurs. Roll a die to determine which one:

\subsubsection*{1-2 Friendly fire incident:} Your unit that is farthest from its Battle Group's Leadership units activates, takes a single Shoot action against its closest friendly unit, gains a Suppression Marker, and then deactivates.

\subsubsection*{3-4 Breakdown:} Your slowest \textit{Vehicle} unit has broken down or run out of fuel; its Move value is permanently reduced to 0.

\subsubsection*{5-6 Earthquake:} The ground shakes with seismic activity or the thunder of weapons fire. During this round, every unit that moves is considered to be in Dangerous Terrain 8+.

\subsubsection*{7-8 Comms are down:} During this round, Leadership units may not issue Orders to units farther than 5`` away or any Call in an airstrike! Orders.

\subsubsection*{9-10 ``It's a trap!'':} Pick an enemy unit in an area of terrain. That area becomes Dangerous Terrain 6+!

If the event rolled does not apply, roll again.

\end{multicols}
\end{document}
